Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-05-15 talk/dls2014/paper/paper.tex
add some missing points

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset (SO)} points to the
private version of an object in all threads\,/\,segments. To then
translate this SO to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
SO. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

x86-CPUs provide a feature called \emph{memory segmentation}. It
performs this translation from a SO to a LA directly in hardware.  We
can use the segment register $\%gs$, which is mostly unused in current
applications. When this register points to a thread's segment start
address, we can instruct the CPU to perform the above translation from
a reference of the form $\%gs{::}SO$ to the right LA on its own. This
process is efficient enough that we can do it on every access to an
object.

In summary, we can use a single SO to reference the same object in all
threads, and it will be translated by the CPU to a LA that always
points to the thread's private version of this object. Thereby,
threads are fully isolated from each other. However, $N$ segments
require $N$-times the memory and modifications on an object need to be
propagated to all segments.

@@

To detect when to privatise pages, we use write barriers before every
write. When the barrier detects that the object is not in a private
page (or any pages that belong to the object), we remap and copy the
pages to the thread's segment. From now on, the translation of
$\%gs{::}SO$ in this particular segment will resolve to the private
version of the object. Note, the SO used to reference the object does
not change during that process.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers
}and already mentioned in section \ref{sub:Setup}. This area can be
seen as a continuous array of bytes that is indexed from the start of
the segment by an object's reference ($SO$) divided by 16 (this
requires objects of at least 16 bytes in size). Instead of just
setting the byte to \lstinline!true!  if the corresponding object was
read, we set it to a \lstinline!read_version!  belonging to the
transaction, which will be incremented on each commit.  Thereby, we
can avoid resetting the bytes to \lstinline!false!  on commit and only
need to do this every 255 transactions. The whole code for the barrier
is easily optimisable for compilers as well as perfectly predictable
for CPUs:

@@

For TM, we first perform a read barrier on the object. We then try to
acquire its write lock. \lstinline!write_locks!  again is a simple
global array of bytes that is indexed with the SO of the object
divided by 16. If we already own the lock, we are done.  If someone
else owns the lock, we will do a write-write contention management
that will abort either us or the current owner of the object.  If we
succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

\section{Experimental Results}

compare some programs between
\begin{itemize}[noitemsep]
\item pypy
\item pypy-jit
\item pypy-stm
\item pypy-stm-jit
\item cpython
\item jython
\end{itemize}

>>>>

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset ($SO$)} points to the
private version of an object in all threads\,/\,segments. To then
translate this $SO$ to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
$SO$. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

x86-CPUs provide a feature called \emph{memory segmentation}. It
performs this translation from a $SO$ to a LA directly in hardware.  We
can use the segment register $\%gs$, which is mostly unused in current
applications. When this register points to a thread's segment start
address, we can instruct the CPU to perform the above translation from
a reference of the form $\%gs{::}SO$ to the right LA on its own. This
process is efficient enough that we can do it on every access to an
object.

In summary, we can use a single $SO$ to reference the same object in all
threads, and it will be translated by the CPU to a LA that always
points to the thread's private version of this object. Thereby,
threads are fully isolated from each other. However, $N$ segments
require $N$-times the memory and modifications on an object need to be
propagated to all segments.

@@

To detect when to privatise pages, we use write barriers before every
write. When the barrier detects that the object is not in a private
page (or any pages that belong to the object), we remap and copy the
pages to the thread's segment. From now on, the translation of
$\%gs{::}SO$ in this particular segment will resolve to the private
version of the object. Note, the $SO$ used to reference the object does
not change during that process.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers}
and already mentioned in section \ref{sub:Setup}.

This area can be seen as a continuous array of bytes that is indexed
from the start of the segment by an object's reference ($SO$) divided
by 16 (this is where the requirement of objects to be of at least 16
bytes in size comes from). Instead of just setting the byte to
\lstinline!true!  if the corresponding object was read, we set it to a
\lstinline!read_version!  belonging to the transaction, which will be
incremented on each commit.  Thereby, we can avoid resetting the bytes
to \lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for CPUs:

@@

For TM, we first perform a read barrier on the object. We then try to
acquire its write lock. \lstinline!write_locks!  again is a simple
global array of bytes that is indexed with the $SO$ of the object
divided by 16. If we already own the lock, we are done.  If someone
else owns the lock, we will do a write-write contention management
that will abort either us or the current owner of the object.  If we
succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

\section{Evaluation}

\subsection{Memory Requirements}

\begin{itemize}
\item stm\_flags per object
\item read markers and other sections
\item private pages
\end{itemize}

maybe some memory usage graph over time

\subsection{Overhead Breakdown}

\begin{itemize}
\item time taken by read \& write barriers
\item time spent committing \& aborting (maybe with different numbers
  of threads)
\item time in GC
\end{itemize}

\subsection{Scaling}

maybe some simple micro benchmarks with adaptable conflict rate

\subsection{Real-World Benchmarks}

more real benchmarks comparing multiple implementations:
\begin{itemize}[noitemsep]
\item pypy
\item pypy-jit
\item pypy-stm
\item pypy-stm-jit
\item cpython
\item jython
\end{itemize}
